Evaluating the Performance and Energy Efficiency of the \textsc{COSMO-ART} model system

### 2014-05-05 paper_enahpc14/Introduction.tex
cleans up latest version; improves format

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid-20$^{th}$  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  determining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  heat-trapping greenhouse  gases and
have  shorter  atmospheric lifetimes,  aerosols  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of solar radiation, thereby exerting a direct radiative forcing; or by
modifying cloud  properties,  cloud  fraction and  surface
albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and  measurement-based  assessments,  large  uncertainties  remain  in
current  estimates  of  aerosol radiative  forcing  \citep{Myhre-2013,
IPCC-2013,  Lee-2013,   Randles-2013,  Rosenfeld-2013,  Sherwood-2013,
Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale modelling system is coupled  with the operational  weather forecast
model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed and used by  a
consortium of European weather centers, as well as utilised in a climate version
by the wider research community.

The extended  atmospheric model \textsc{Cosmo-art}  is com\-put\-ationally
much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
additional  tracers and  processes have  to be  considered.   Thus the
model  is currently  severely limited  in terms  of applicability  and
expensive in terms of energy consumption.  Although \textsc{Cosmo} has
recently been ported to GPUs \citep{Gysi-2014, Lapillonne-2014} within
the framework of the  High Performance and High Productivity Computing
(HP2C) Initiative  \citep{HP2C} to optimise it  for computational and
energy efficiency,  significant investments in ART  are still required
to take it  to a similar level.  The  efficiency of \textsc{Cosmo-art}
is being  addressed in the  EU Exa2Green project  \citep{EXA2GREEN} to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
introduce \textsc{Cosmo-art}  and specify its  technical setup related
to  the investigated  performance and  energy evaluation  methods.  In
Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking study.  Sec.~\ref{sec:4}  presents performance and energy
requirements  of the  baseline on  these architectures  and highlights
areas where improvements will be necessary for the subsequent baseline
refactoring.   Finally,  we  conclude in Sec.~\ref{concl} with some 
implications for the Exa2Green project and give an outlook for future
research.

>>>>

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate changes  since  the mid-20$^{th}$  Century, one  factor
overwhelmingly  affects the  uncertainty in  determining human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not  considered heat-trapping greenhouse  gases and
have shorter atmospheric  lifetimes, aerosols significantly modify the
global   radiation   budget   \cite{IPCC-2013}.    Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of solar radiation, thereby exerting a direct radiative forcing; or by
modifying cloud properties, cloud fraction and surface albedo, causing
a negative indirect  radiative forcing.  Despite considerable progress
in  global aerosol  modelling  \cite{Mann-2013} and  measurement-based
assessments,  large  uncertainties  remain  in  current  estimates  of
aerosol  radiative   forcing  \cite{Myhre-2013,  IPCC-2013,  Lee-2013,
Sherwood-2013, Stier-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
scale  modelling  system  is  coupled  with  the  operational  weather
forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed
and  used by  a consortium  of European  weather centers,  as  well as
utilized in a climate version by the wider research community.

The extended  atmospheric model \cosmoart  is com\-put\-ationally much
more demanding than \textsc{Cosmo}  since a large number of additional
tracers  and processes  have  to  be considered.   Thus  the model  is
currently severely limited in  terms of applicability and expensive in
terms  of energy  consumption.  Although  \textsc{Cosmo}  has recently
been  ported  to  GPUs  \cite{Gysi-2014, Lapillonne-2014}  within  the
framework  of the  High  Performance and  High Productivity  Computing
(HP2C)  Initiative \cite{HP2C}  to optimize  it for  computational and
energy efficiency,  significant investments in ART  are still required
to take it  to a similar level.  The efficiency  of \cosmoart is being
addressed in  the EU Exa2Green  project \cite{EXA2GREEN} to  deliver a
prototype code, which  provides an energy efficiency of  at least five
times  the baseline  value.  Such  an implementation  would  allow the
community to  investigate critical questions at  higher resolution and
over longer periods, at reduced cost to the environment.

This work is  organized as follows: in Section~\ref{sec:1}  we give an
overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly
introduce  \cosmoart and specify  its technical  setup related  to the
investigated   performance   and   energy  evaluation   methods.    In
Section~\ref{sec:3}  we  describe  HPC  platforms,  power  measurement
equipments   and  software  environment   employed  to   conduct  this
benchmarking  study.   Section~\ref{sec:4}  presents  performance  and
energy  requirements  of  the  baseline  on  these  architectures  and
highlights  areas  where  improvements   will  be  necessary  for  the
subsequent   baseline   refactoring.     Finally,   we   conclude   in
Section~\ref{concl} with  some implications for  the Exa2Green project
and give an outlook for future research.

### 2014-05-05 paper_enahpc14/Sec4:Experimental-results.tex
Adding missing figure
cleans up latest version; improves format

For the experiments on \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings. Specifically, the OpenMPI 1.6.5 library, 
features two operation modes, blocking and polling, which can be selected before \cosmoart
is launched. In the polling configuration (the default mode) ``MPI'' continually polls
the network interface to check for the completion of an event (e.g. send or receive).
This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
attains the best performance, possibly at the cost a higher energy usage. In the blocking
mode, the CPU is yield to other processes/threads if there are no incoming messages~\cite{PDP12}.

@@

Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
\monch's Isola E1 Rack  2 and \pilat' Isola HD total
power measurements for  1-day or 2-day simulations. On  the Intel Ivy
Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
was issued  only twice due  to usage restrictions. As  time resolution
was set to one update every 5 minutes for power sampling, the averaged
power dissipation was computed by considering 6 values for each single
run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
1-day simulation  was issued  four times and  a 2-day run  only once.
Similarly, the averaged power dissipation was computed by considering 4
values  for  each  single  1-day  run  and 9  values  for  the  2-days
run. Corresponding results are gathered in Table~\ref{tab:3}.

@@

\begin{table}[htbf]
  \centering
    \caption{Averaged power dissipation (W) of the platforms.}
    \label{tab:3}
    \begin{tabular}{cccc}
      \hline\noalign{\smallskip}
      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}

@@

Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. The corresponding results for the averaged total power are also shown in Table~\ref{tab:3}.  
As expected, the average total power dissipation by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy. This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set. From the point of view of the execution time, there is a slight reduction for the blocking MPI mode. One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the execution time, however, applications with extensive computing, as e.g. \cosmoart, may show better performance with the blocking modes. This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.

@@

\begin{figure}[htbf]

  \centering
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
  \vspace{-1cm}
  \caption{Mean time-to-solution and energy-to-solution on
    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
    an improvement for blocking policy in both TTS and ETS.}
  \label{fig:4}
\end{figure}

\begin{figure*}[ht]
  \centering
  \hspace{0.6cm}
  \scalebox{0.64}{\input{Figs/24hour.pstex_t}}
  \caption{24 hours simulation power trace using the polling and blocking MPI policies on \tinto.}
  \label{fig:9}
\end{figure*}

\begin{figure*}[ht]
  \centering
  \scriptsize
  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy\\
  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy\\
  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy\\
  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy\\
  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
& &  & &  & &  \\[-0.15cm]
Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{}  & & 
Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{}  & & 
Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{}  & & 
Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,255,  0}}p{0.4cm}}{}  \\
& &  & &  & &  \\[-0.15cm]
Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{}  & &
Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{}  & &
Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  \\
%I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
%
%Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
\end{tabular}
  \caption{Power-performance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
  \label{fig:11}
\end{figure*}

@@

% For all the experiments we also analyze the contribution of the MPI library to the energy 
% consumption and how the use of the blocking versus polling message-passing policies can 
% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
% launched. In the polling configuration (the default mode) ``MPI'' continually polls
% the network interface to check for the completion of an event (e.g. send or receive).
% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
% mode, the CPU is yield to other processes/threads if there are no incoming messages.

In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a 1-day simulation of \cosmoart. These profiles, produced with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}. The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours. We relate these variations due to the different chemical reactions, aerosols, radiations and clouds formation processes, computed by ART, taking place during daylight hours but not during night. In this case the peak power consumption is about 3821\,W. Looking at the power profile using the blocking MPI policy, we observe that the 1-hour pattern is repeated along the 24-hours, favoring the reduction of power during night hours in a more noticeable way than that with the polling policy. The peak power dissipation during day hours of while the blocking MPI policy is leveraged is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode. Also, we notice systematic power drops for each simulated hour. We associate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage. Our final observation is the reduction of the TTS by 2\,\% with respect to the blocking MPI policy.

To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces to allow us correlate the type of tasks being executed with the total power dissipation along the time (see Figure~\ref{fig:11}). For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a 1-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB). To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and  midnight (from 23h to 24h), in which \cosmoart perform different internal computations. At the same time we leverage the two already stated MPI policies: polling and blocking.

All the traces in Figure~\ref{fig:11} are divided in different computation phases. In first phase all the processes perform group communications and execute MPI-receive primitives (e.g. \texttt{MPI\_Recv}). The second one performs computation bursts, combined in between with group communications, blocking sends, immediate receives and MPI-wait primitives are executed. Finally, group communications interlaced with computation and a significant synchronization phase are performed. From the performance point of view, one can notice a reduction of 2\,\% in the TTS that blocking MPI mode leads over the polling MPI policy. On the other hand an averaged reduction of power dissipation of 2\,\% is attained with the blocking MPI policy only in routines that actually perform synchronizations, group communications and MPI-receive/-wait primitives. Thus, the reduction of the power-energy pair directly depends on the percentage of time in which blocking MPI operations are executed. For the 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked. Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute an approximation of the power savings, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (around 2\,\%). However, we expect higher power-energy savings if future architectures tend proportional computing~\citep{Barroso:2007:CEC:1339817.1339894}.

% \begin{figure*}[htbf]
%   \centering
%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
%   \caption{blabla}
%   \label{fig:8}
% \end{figure*}

>>>>

For  the   experiments  on  \tinto  platform,  we   also  analyze  the
contribution of the MPI library  to the energy consumption and how the
use  of  the  blocking  versus polling  message-passing  policies  can
potentially  render energy  savings. Specifically,  the  OpenMPI 1.6.5
library, features two operation modes, blocking and polling, which can
be selected before \cosmoart is launched. In the polling configuration
(the default mode) ``MPI''  continually polls the network interface to
check for  the completion  of an event  (e.g. send or  receive).  This
usually yields  low latency  but high CPU  utilization. Thus,  one can
expect that  this mode attains  the best performance, possibly  at the
cost a higher energy usage. In  the blocking mode, the CPU is yield to
other     processes/threads    if     there     are    no     incoming
messages~\cite{Castillo-2012}.

@@

Figures~\ref{fig:1} and \ref{fig:2}  account respectively for \monch's
Isola  E1 Rack 2  and \pilat'  Isola HD  total power  measurements for
1-day or 2-day  simulations. On the Intel Ivy  Bridge EP based cluster
(i.e.   \monch), the  1-day simulation  was issued  only twice  due to
usage restrictions. As  time resolution was set to  one update every 5
minutes  for  power  sampling,  the  averaged  power  dissipation  was
computed by  considering 6 values for  each single run.   On the Intel
Xeon E5 based cluster (i.e.   \pilat), the 1-day simulation was issued
four times and  a 2-day run only once.   Similarly, the averaged power
dissipation was computed by considering 4 values for each single 1-day
run  and  9 values  for  the  2-days  run. Corresponding  results  are
gathered in Table~\ref{tab:1}.

@@

\begin{table}[htbf]
  \centering
    \caption{Averaged power dissipation (W) of the platforms.}
    \label{tab:1}
    \begin{tabular}{cccc}
      \hline\noalign{\smallskip}
      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & 
      \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}} \\
      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}

@@

Averaged TTS and ETS results  over 10 repetitions of 24h simulation on
\tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
the  averaged total  power are  also shown  in  Table~\ref{tab:1}.  As
expected, the average total power dissipation by \cosmoart while using
the blocking MPI policy is about  2\,\% less than that consumed by the
polling  policy. This is  due to  the power-friendly  (blocking) waits
performed  by the  CPU cores  while using  the blocking  MPI  mode, in
contrast to  the polling  waits that take  place when the  polling MPI
mode is set. From the point of  view of the execution time, there is a
slight reduction  for the blocking MPI  mode. One can  expect that the
costs of the interruptions used in the blocking MPI policy can lead to
an  increase  of  the   execution  time,  however,  applications  with
extensive computing,  as e.g.  \cosmoart, may show  better performance
with the blocking  modes. This effect works in  favor with our results
since the energy consumed is also reduced (2\,\%) just by changing the
behavior of the OpenMPI library  to the blocking policy, i.e, with the
\texttt{--mca    mpi\_yield\_when\_idle}   parameter   set    in   the
\texttt{mpirun} command.

@@

\begin{figure}[htbf]
  \centering
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
  \vspace{-1cm}
  \caption{Mean time-to-solution and  energy-to-solution on \tinto for
    a 24h simulation using the  polling and blocking MPI policies. The
    95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
    samples) clearly  indicates an improvement for  blocking policy in
    both TTS and ETS.}
  \label{fig:4}
\end{figure}

\begin{figure*}[ht]
  \centering
  \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
  \caption{24  hours  simulation power  trace  using  the polling  and
    blocking MPI policies on \tinto.}
  \label{fig:5}
\end{figure*}

\begin{figure*}[ht]
  \centering
  \scriptsize
  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy \\
  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}} \\
  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy \\
  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}} \\
  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy \\
  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}} \\
  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy \\
  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}} \\
  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
& &  & &  & &  \\[-0.15cm]
Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{} & & 
Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{} & & 
Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{} & & 
Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,255,  0}}p{0.4cm}}{} \\
& &  & &  & &  \\[-0.15cm]
Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{} & &
Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{} & & 
Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{} & &
Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{} \\
%I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{} & &
%
%Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{} \\
\end{tabular}
  \caption{Power-performance   traces  during   midday   and  midnight
    leveraging the polling and blocking MPI policies on \tinto.}
  \label{fig:6}
\end{figure*}

@@

In order to obtain an initial overview of the power dissipation of the
system  model,  we  obtained  power  profiles  leveraging  our  \pmlib
power-measurement    framework   over    a    1-day   simulation    of
\cosmoart. These profiles, produced with both polling and blocking MPI
policies,  are depicted in  Figure~\ref{fig:5}. The  first observation
for the polling MPI policy is that the first and last hours of the day
dissipate slightly less power than  the mid-day hours. We relate these
variations  due   to  the  different   chemical  reactions,  aerosols,
radiations  and clouds  formation processes,  computed by  ART, taking
place during  daylight hours  but not during  night. In this  case the
peak power consumption is about  3821\,W. Looking at the power profile
using the blocking  MPI policy, we observe that  the 1-hour pattern is
repeated along  the 24-hours, favoring  the reduction of  power during
night  hours in  a  more noticeable  way  than that  with the  polling
policy.  The peak  power dissipation  during  day hours  of while  the
blocking  MPI policy  is leveraged  is  about 3768\,W,  which means  a
reduction of around  50\,W with respect to the  polling mode. Also, we
notice systematic  power drops for  each simulated hour.  We associate
them  to periods  in  which  the processes  were  blocked waiting  for
incoming messages  or synchronizations, thus  yielding the CPU  to low
usage. Our final observation is the reduction of the TTS by 2\,\% with
respect to the blocking MPI policy.

To  gain  more  insights   about  the  power-performance  behavior  of
\cosmoart,  we  obtained power-per\-for\-man\-ce  traces  to allow  us
correlate  the type  of  tasks  being executed  with  the total  power
dissipation  along  the   time  (see  Figure~\ref{fig:6}).   For  this
experiment we  limit the simulation  time to 1 hour  because \emph{i})
the  pattern of  the computation  performed by  \cosmoart  is repeated
hourly and, \emph{ii}) the weight  of a 1-day simulation trace can not
be easily handled using our  machines and memory available (each trace
file was about 45\,GB). To  face the problem we shrink the simulations
to 1 hour  and test a time  frame during midday (from 11h  to 12h) and
midnight  (from 23h  to  24h), in  which  \cosmoart perform  different
internal computations.  At  the same time we leverage  the two already
stated MPI policies: polling and blocking.

All  the  traces  in   Figure~\ref{fig:6}  are  divided  in  different
computation  phases. In first  phase all  the processes  perform group
communications      and       execute      MPI-receive      primitives
(e.g. \texttt{MPI\_Recv}). The second one performs computation bursts,
combined  in  between   with  group  communications,  blocking  sends,
immediate  receives and  MPI-wait primitives  are  executed.  Finally,
group  communications interlaced  with computation  and  a significant
synchronization  phase are  performed. From  the performance  point of
view, one can notice a reduction of 2\,\% in the TTS that blocking MPI
mode leads over the polling MPI  policy. On the other hand an averaged
reduction of power dissipation of  2\,\% is attained with the blocking
MPI policy  only in  routines that actually  perform synchronizations,
group  communications  and  MPI-receive/-wait primitives.   Thus,  the
reduction of the power-energy  pair directly depends on the percentage
of time in which blocking MPI  operations are executed. For the 1 hour
simulation traces, an average of 50\,\% of the time the processes were
blocked.  Taking  into account that  the total power  reduction during
blocking phases  is about 5\,\%,  one can compute an  approximation of
the  power savings,  i.e., $50\,\%  \times  5\,\% =  2.5\%$, which  is
consistent  power  reduction  obtained  for  the  full-day  simulation
(around  2\,\%). However,  we  expect higher  power-energy savings  if
future architectures tend proportional computing~\citep{Barroso-2007}.

% \begin{figure*}[htbf]
%   \centering
%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
%   \caption{blabla}
%   \label{fig:7}
% \end{figure*}
